Demo: Counterpoint by Construction

### 2019-06-28 doc/farm/farm-abstract.tex
camera-ready

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,10pt,screen]{acmart}
\settopmatter{printfolios=true, printccs=true, printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

\acmConference[FARM 2019]{The ACM SIGPLAN International Workshop on Functional Art, Music, Modelling and Design}{August, 2019}{Berlin, Germany}
\acmYear{2019}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

@@

To analyze and synthesize tonal music, researchers have
attempted to encode these rules into programming languages.
Functional programming languages seem
ideally suited for this task, and in particular, those with a static
type system can further guarantee that \emph{well-typed music 
does not sound wrong}.
In the past decade, Haskell has been extensively used to encode
the rules of harmony \citep{fmmh, hihseufha, faamh, hafha, fghm}
as well as counterpoint \citep{szamozvancev2017well}.
An interesting observation is that, many of the existing encodings
rely on some form of \emph{dependent types}, i.e.,  
types that depend on terms. 
Since Haskell is not a dependently typed language, one has to
use language extensions, such as GADTs \citep{cheney2002lightweight} 
and singleton types \citep{eisenberg2013dependently},
to simulate dependencies.
While this allows encoding a wide class of music rules, it can
require duplicating code to reflect terms into 
types, making the implementation less elegant 
\citep{monnier2010singleton}.
This motivates us to explore music programming in a language
with intrinsic support for dependent types.

@@

The goal of Music Tools is to provide a core collection of types and functions
that can be used to easily build applications to analyze and synthesize music.
It is intended to evolve into a dependently-typed replacement for Euterpea
\citep{hudak2018haskell} and borrows many ideas from that library.
It is currently restricted to the chromatic scale for simplicity, and like
Euterpea is designed to work well with MIDI.

@@

To play music on synthesizers, one can convert it to MIDI by calling 
Haskell MIDI libraries via Agda's Haskell FFI. 
Details can be found in the github repository.

@@

The last requirement is that the music must end with a cadence,
which is a final motion from the 2nd or 7th degree to the tonic 
(1st degree). 
We impose this requirement by declaring two cadence constructors 
as the base cases of counterpoint (note that the final interval of
the cadence is always \texttt{(p , per8)} and is thus not explicitly
specified).
Thus, we arrive at the following datatype for well-typed counterpoint\footnote{
For readability, we have omitted explicit conversions 
from \texttt{PitchInterval} (which ensures the interval is not dissonant) 
to the general \texttt{Interval}.}.

@@

\noindent Observe that \texttt{motionOk} is an implicit argument of 
the \texttt{\_::\_} constructor. 
The argument can be resolved automatically by the type checker,
hence there is no need to manually supply this proof.

@@

We intend to expand the Music Tools library to include
richer functionality for analysis and synthesis. We are
particularly interested in expressing the work done in Haskell
on functional harmony in the library, to see just how much
dependent types can simplify the representation.

@@

For counterpoint specifically, we plan to represent higher species 
counterpoint by extending the rules, 
and to explore automatic generation of species counterpoint.
It would be interesting to compare our correct-by-construction counterpoint
with that created by machine learning \citep{CounterpointByConvolution}, 
which does not have correctness guarantees.

%% Acknowledgments
\begin{acks}                            %% contents suppressed with 'anonymous'
 The authors would like to thank the participants of the Tokyo Agda 
 Implementors' Meeting, especially Ulf Norell and Jesper Cockx,
 for many helpful suggestions that improved our Agda code.
 We also thank the anonymous reviewers for their thoughtful feedback
 which improved our presentation.
\end{acks}

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,10pt,screen]{acmart}
\settopmatter{printfolios=true, printccs=true, printacmref=true}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%%% The following is specific to FARM '19 and the paper
%%% 'Demo: Counterpoint by Construction'
%%% by Youyou Cong and John Leo.
%%%
\setcopyright{rightsretained}
\acmPrice{}
\acmDOI{10.1145/3331543.3342578}
\acmYear{2019}
\copyrightyear{2019}
\acmISBN{978-1-4503-6811-7/19/08}
\acmConference[FARM '19]{Proceedings of the 7th ACM SIGPLAN International Workshop on Functional Art, Music, Modeling, and Design}{August 23, 2019}{Berlin, Germany}
\acmBooktitle{Proceedings of the 7th ACM SIGPLAN International Workshop on Functional Art, Music, Modeling, and Design (FARM '19), August 23, 2019, Berlin, Germany}

@@

To analyze and synthesize tonal music, researchers have
attempted to encode these rules into programming languages.
Functional programming languages seem
ideally suited for this task, and in particular, those with a static
type system can further guarantee that \emph{well-typed music 
does not sound wrong}.
In the past decade, Haskell has been a popular choice for encoding
the rules of harmony \citep{fmmh, hihseufha, faamh, hafha, fghm}
as well as counterpoint \citep{szamozvancev2017well}.
An interesting observation is that, many of the existing encodings
rely on some form of \emph{dependent types}, i.e.,  
types that depend on terms. 
While Haskell is not a dependently typed language, it has various 
language extensions (such as GADTs \citep{cheney2002lightweight} 
and singleton types \citep{eisenberg2013dependently}) for 
simulating dependent types, and as shown by previous studies,
the extended type system is expressive enough for 
a wide class of music applications.
On the other hand, the need for simulating dependent types 
can also result in code duplication, making the implementation 
less elegant \citep{monnier2010singleton}.
This motivates us to explore music programming in a language
with intrinsic support for dependent types.

@@

The goal of Music Tools is to provide a core collection of types and functions
that can be used to easily build applications to analyze and synthesize music.
It is intended to evolve into a dependently-typed replacement for Euterpea
\citep{hudak2018haskell}, and borrows many ideas from that library.
Currently, Music Tools is restricted to the chromatic scale for simplicity, 
and like Euterpea is designed to work well with MIDI.

@@

To play music on synthesizers, one can convert it to MIDI by calling 
Haskell MIDI libraries via Agda's Haskell FFI. 
Details can be found in the GitHub repository.

@@

The last requirement is that the music must end with a cadence,
which is a final motion from the 2nd or 7th degree to the tonic 
(1st degree). 
We impose this requirement by declaring two cadence constructors 
as the base cases of counterpoint (note that the final interval of
the cadence is always \texttt{(p\ ,\ per8)} and hence is not explicitly
specified).
Thus, we arrive at the following datatype for well-typed counterpoint\footnote{
For readability, we have omitted explicit conversions 
from \texttt{PitchInterval} (which ensures the interval is not dissonant) 
to the general \texttt{Interval}.}.

@@

\noindent Observe that \texttt{motionOk} is an implicit argument of 
the \texttt{\_::\_} constructor, here denoted by the curly braces. 
The argument can be resolved automatically by the type checker,
hence there is no need to manually supply this proof.

@@

We intend to expand the Music Tools library to include
richer functionality for analysis and synthesis. We are
particularly interested in expressing the work done in Haskell
on functional harmony in the library, to see how much
dependent types can simplify the representation.

@@

For counterpoint specifically, we intend to represent higher species 
counterpoint by extending the rules, 
and to explore automatic generation of species counterpoint.
It would be interesting to compare our correct-by-construction counterpoint
with that created by machine learning \citep{CounterpointByConvolution}, 
which does not have correctness guarantees.

%% Acknowledgments
\begin{acks}                            %% contents suppressed with 'anonymous'
 The authors would like to thank the participants of the Tokyo Agda 
 Implementors' Meeting, especially Ulf Norell and Jesper Cockx,
 for many helpful suggestions that improved our Agda code.
 We also thank the anonymous reviewers for their thoughtful feedback,
 which improved our presentation.
\end{acks}
